---
title: "JaxBatch1_ReDo"
author: "Mike Martinez"
date: "2023-11-14"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, set working directory}
setwd("/Users/mikemartinez/Desktop/EOLO/EOLO/JaxBatch1/")
library(imcRtools)
library(cytomapper)
library(pheatmap)
```

Read in the steinbock output into a `SpatialExperiment` object
```{r read steinbock}
# Read in the steinbock output folder
spe <- read_steinbock("/Users/mikemartinez/Desktop/EOLO/EOLO/JaxBatch1/steinbock/")

# OR....

#Read in the spe object
spe <- readRDS("/Users/mikemartinez/Desktop/EOLO/EOLO/JaxBatch1/RDS_objects/Kmeans_clustered_spe.rds")
```

```{r Formatting spatial experiment object}
# Format sample_id, colnames, and rownames
spe$sample_id <- sub("12222021_Rosenberg_TMA_1_", "ROI_", spe$sample_id)
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
rownames(spe) <- c("aSMA", "Vimentin", "CD14", "CD31", "CD56", "CD20", "CD45RA", "CD11c", "CD4", "Epcam_ECadherin", "MHC_II", "CD8a", "CD68", "CD45", "CD15",
                   "FOXP3", "Tryptase", "GZMB", "Ki67", "Collagen_I", "CD3", "MHC_I", "CD45RO", "Pan_Keratin", "DNA1", "DNA3")

# Specify the channels to be used in downstream analysis. Since DNA1 and DNA3 are just DNA intercolators, we will ignore them moving forward
rowData(spe)$use_channel <- !grepl("DNA1|DNA3", rownames(spe))
```

```{r transform counts, de-select nuclear markers for downstream analysis}
library(dittoSeq)

# Plot the nuclear marker expression before asinh transforming
dittoRidgePlot(spe, var = c("DNA1", "DNA3"), group.by = "sample_id", assay = "counts") +
  ggtitle("Nuclear Markers: Before asinh Transformation")

# asinh transform
assay(spe, "exprs") <- asinh(counts(spe)/1)
dittoRidgePlot(spe, var = c("DNA1", "DNA3"), group.by = "sample_id", assay = "exprs") +
  ggtitle("Nuclear Markers: After asinh Transformation")

# All markers
markers <- unique(rownames(spe))

#----- Run this to check the cytoplasmic markers
# Iterate through each cytoplasmic marker and plot before and after asinh transformation
for (marker in markers) {
  before <- dittoRidgePlot(spe, var = marker, group.by = "sample_id", assay = "counts") +
              ggtitle(paste(marker, "Before asinh transformation", sep = " "))
  #ggsave(paste(marker, "Before_asinh_transformation.pdf", sep = "_"), before)
  
  after <- dittoRidgePlot(spe, var = marker, group.by = "sample_id", assay = "exprs") +
            ggtitle(paste(marker, "After asinh transformation", sep = " "))
  #ggsave(paste(marker, "After_asinh_transformation.pdf", sep = "_"), after)
}
```

```{r load in images and masks}
library(cytomapper)

# Read in images and masks
images <- loadImages("/Users/mikemartinez/Desktop/EOLO/EOLO/JaxBatch1/steinbock/img/")
masks <- loadImages("/Users/mikemartinez/Desktop/EOLO/EOLO/JaxBatch1/steinbock/masks_deepcell/", as.is = TRUE)

# Set channel names for images...should match spe
channelNames(images) <- rownames(spe)
names(images) <- sub("12222021_Rosenberg_TMA_1_", "ROI_", names(images))
names(masks) <- sub("12222021_Rosenberg_TMA_1_", "ROI_", names(masks))

# Set image names
mcols(images) <- mcols(masks) <- DataFrame(sample_id = names(images))
```

```{r metadata}
library(RColorBrewer)

# Read in the metadata
meta <- read.csv("/Users/mikemartinez/Desktop/EOLO/EOLO/JaxBatch1/Metadata/JaxBatch1_Meta.csv", header = TRUE, sep = ",")

# Get a vector of sample IDs
Sample_ID <- names(images)

# Assign metadata for patient_id
spe$patient_id <- meta$Patient_ID[match(spe$sample_id, meta$Sample_ID)]
spe$patient_id <- factor(spe$patient_id)
patient <- c("2955", "925", "5057", "7341")

# Assign metadata for indication
spe$indication <- meta$Indication[match(spe$sample_id, meta$Sample_ID)]
indication <- c("Early", "Early", "Late", "Late", "Early", "Early", "Late", "Late")

# Assign metadata for type
spe$type <- meta$Type[match(spe$sample_id, meta$Sample_ID)]
Type <- c("NonTumor", "Tumor")

#Assign the metadata to the images and masks as well
mcols(images) <- mcols(masks) <- DataFrame(sample_id = names(images),
                                           patient_id = patient,
                                           indication = indication,
                                           type = Type)

#Initialize an empty list to store color vectors which will be stored in the spe metadata slot
color_vectors <- list()

# patient_id color vector
patient_id <- setNames(brewer.pal(length(unique(spe$patient_id)), name = "RdBu"),
                                  unique(spe$patient_id))
# sample_id color vector
sample_id <- setNames(brewer.pal(length(unique(spe$sample_id)), name = "Paired"),
                                 unique(spe$sample_id))
# type color vector 
type <- c("NonTumor" = "green",
          "Tumor" = "orange")

# indication color vector
indication <- c("Late" = "red",
                "Early" = "blue")

# Append color vectors to to color_vectors list
color_vectors$patient_id <- patient_id
color_vectors$sample_id <- sample_id
color_vectors$indication <- indication
color_vectors$type <- type

# Assign color_vectors list to metadata slot of spe
metadata(spe)$color_vectors <- color_vectors
```


```{r segmentation quality control 1: Masks on ROIs}
library(dittoSeq)
library(viridis)
library(cytomapper)
library(gridGraphics)
library(cowplot)

# Seed seed for reproducibility
set.seed(03061999)

# Randomly sample 2 ROIs out of the 8 ROIs
img_ids <- sample(seq_along(images), 2)
cur_images <- images[img_ids]
cur_masks <- masks[img_ids]

# 2 step normalization scheme. Normalize across images --> normalize within images.
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0,0.2))

# Look at the segmentation using a nuclear marker and a few cytoplasmic markers
segmentQC <- plotPixels(image = cur_images,
           mask = cur_masks,
           object = spe,
           cell_id = "ObjectNumber", 
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA1","Epcam_ECadherin", "CD3", "CD20"),
           colour = list(DNA1 = c("black", "blue"),
                         Epcam_ECadherin = c("black", "green"),
                         CD3 = c("black", "red"),
                         CD20 = c("black", "orange")),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           return_plot = TRUE)

# Plot and save
QC <- ggdraw(segmentQC$plot, clip = "on")
ggsave("segmentationQC.png", QC, width =12, height = 8)
```

```{r segmentation quality control 2: Single-cell patterns}
library(dittoSeq)
library(viridis)

# Sub-sample the cells
cur_cells <- sample(seq_len(ncol(spe)), 2000)

# Plot heatmap
scMarkerPatterns <- dittoHeatmap(spe[,cur_cells],
             assay = "exprs",
             cluster_cols = FALSE,
             scale = "none",
             annot.by = c("indication", "type"),
             annotation_colors = list(indication = metadata(spe)$color_vectors$indication,
                                      type = metadata(spe)$color_vectors$type),
             heatmap.colors = viridis(100),
             main = "Single-cell Expression Patterns Across Samples")
ggsave("MarkerPatterns_acrossPatients_ROIs.pdf", scMarkerPatterns, width = 10, height = 8)
```

```{r segmentation quality control 3: Mean marker expression per image}
library(scuttle)

# Aggregate cells and calculate the mean expression
image_mean <- aggregateAcrossCells(spe,
                                   ids = spe$sample_id,
                                   statistics = "mean",
                                   use.assay.type = "counts")
assay(image_mean, "exprs") <- asinh(counts(image_mean))

# Plot heatmap 
markerExpression <- dittoHeatmap(image_mean,
             assay = "exprs",
             cluster_cols = FALSE,
             scale = "none",
             heatmap.colors = viridis(100),
             annot.by = c("patient_id", "sample_id"),
             annotation_colors = list(patient_id = metadata(spe)$color_vectors$patient_id,
                                      sample_id = metadata(spe)$color_vectors$sample_id),
             show_colnames = TRUE,
             main = "Marker Expression Levels Across All Images")
ggsave("PerImage_MarkerExpression.pdf", markerExpression, width = 10, height = 8)

```

#-----For below: 65536 is a number representative of the number of intensity levels in a 16-bit images (which Steinbock outputs)
16-bit images have 2^16 = 65536 intensity levels

```{r Image quality control 1A: Signal to noise ratio}
library(tidyverse)
library(ggrepel)
library(EBImage)

# Calculate the signal to noise ratio
cur_snr <- lapply(names(images), function(x){
  img <- images[[x]]
  mat <- apply(img, 3, function(ch){
    # Otsu threshold
    thres <- otsu(ch, range = c(min(ch), max(ch)), levels = 65536)
    # Signal-to-noise ratio
    snr <- mean(ch[ch > thres]) / mean(ch[ch <= thres])
    # Signal intensity
    ps <- mean(ch[ch > thres])
    
    return(c(snr = snr, ps = ps))
  })
  t(mat) %>% as.data.frame() %>%
    mutate(image = x, 
           marker = colnames(mat)) %>%
    pivot_longer(cols = c(snr, ps))
})

cur_snr <- do.call(rbind, cur_snr)

# Plot SNR
cur_snr %>%
  group_by(marker, name) %>%
  summarize(log_mean = log2(mean(value))) %>%
  pivot_wider(names_from = name, values_from = log_mean) %>%
  ggplot() +
  geom_point(aes(ps, snr)) +
  geom_label_repel(aes(ps, snr, label = marker)) +
  theme_minimal(base_size = 15) + ylab("Signal-to-noise ratio [log2]") +
  xlab("Signal intensity [log2]")
```

```{r Image quality control 1B: Signal to noise ratio}
cur_snr <- cur_snr %>%
  pivot_wider(names_from = name, values_from = value) %>%
  filter(ps > 2) %>%
  pivot_longer(cols = c(snr, ps))

# Plot
cur_snr %>%
  group_by(marker, name) %>%
  summarize(log_mean = log2(mean(value))) %>%
  pivot_wider(names_from = name, values_from = log_mean) %>%
  ggplot() +
  geom_point(aes(ps, snr)) +
  geom_label_repel(aes(ps, snr, label = marker)) +
  theme_minimal(base_size = 15) + ylab("Signal-to-noise ratio [log2]") +
  xlab("Signal Intensity [log2]")

```

```{r Image quality control 2: Cell Area}

# Calculate the number of pixels covered by cells
cell_density <- colData(spe) %>%
  as.data.frame() %>%
  group_by(sample_id) %>%
  
  # Divide the total number of pixels by the number of pixels covered by cells
  summarize(cell_area = sum(area),
            no_pixels = mean(width_px) * mean(height_px)) %>%
  mutate(covered_area = cell_area / no_pixels)

# Plot
cellCoverage <- ggplot(cell_density) +
  geom_point(aes(reorder(sample_id, covered_area), covered_area)) +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 15)) +
  ylim(c(0,1)) +
  ylab("% Covered area") + xlab("")
ggsave("Percent_Covered_Area_per_Image.pdf", cellCoverage, width = 10, height = 8)

```

```{r Image quality control 3: SNR at cell level using Gaussian Mixture Models}
library(mclust)

# Set seed for reproducibility
set.seed(03061999)

mat <- sapply(seq_len(nrow(spe)), function(x){
  cur_exprs <- assay(spe, "exprs")[x,]
  cur_counts <- assay(spe, "counts")[x,]
  
  cur_model <- Mclust(cur_exprs, G = 2)
  mean1 <- mean(cur_counts[cur_model$classification == 1])
  mean2 <- mean(cur_counts[cur_model$classification == 2])
  
  signal <- ifelse(mean1 > mean2, mean1, mean2)
  noise <- ifelse(mean1 > mean2, mean2, mean1)
  
  return(c(snr = signal/noise, ps = signal))
})

cur_snr <- t(mat) %>% as.data.frame() %>%
  mutate(marker = rownames(spe))

cur_snr %>% ggplot() +
  geom_point(aes(log2(ps), log2(snr))) +
  geom_label_repel(aes(log2(ps), log2(snr), label = marker)) +
  theme_minimal(base_size = 15) + ylab("Signal-to-noise ratio [log2]")+
  xlab("Signal intensity [log2]")

```

```{r Image quality control 4: Cell Sizes Across All ROIs}

# Calculate the distribution of cell sizes across each ROI
dist <- dittoPlot(spe, var = "area",
          group.by = "sample_id",
          plots = "boxplot") +
  ylab("Cell area") + xlab("") +
  ggtitle("Distribution of cell sizes across images")
ggsave("CellSize_Distributions_Across_Images.pdf", dist, width = 10, height = 8)

# Summarize the cell areas
summary(spe$area)

```

```{r Image quality control 5: Removing Small Cells}
sum(spe$area < 5)
spe <- spe[,spe$area >= 5]
```

```{r Image quality control 6: Cell Density and Marker Expression Across Patients}

# Calculate the cell density
cell_density <- colData(spe) %>%
  as.data.frame() %>%
  group_by(sample_id) %>%
  summarize(cell_count = n(),
            no_pixels = mean(width_px) * mean(height_px)) %>%
  mutate(cells_per_mm2 = cell_count/(no_pixels/1000000))

# Plot
cellDensity <- ggplot(cell_density) +
  geom_point(aes(reorder(sample_id, cells_per_mm2), cells_per_mm2)) +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
  ylab("Cells per mm2") + xlab("") + ggtitle("Cell Density per Image")
ggsave("CellDensity_Per_Image.pdf", cellDensity, width = 10, height = 8)

# Plot marker expression for each patient
multiDitto <- multi_dittoPlot(spe, vars = rownames(spe)[rowData(spe)$use_channel],
                group.by = "patient_id", plots = "ridgeplot",
                assay = "exprs",
                legend.show = FALSE,
                color.panel = metadata(spe)$color_vectors$patient_id)
ggsave("MultiDitto.pdf", multiDitto, width = 12, height = 16)
```

```{r Analysis 1A: Dimensional Reduction}
library(scater)

# Set seed for reproducibility
set.seed(03061999)

# Run UMAP and tSNE
spe <- runUMAP(spe, exprs_values = "exprs")
#spe <- runTSNE(spe, exprs_values = "exprs")

# Dimensional reductions are stored in the reducedDims slot of spe
reducedDims(spe)

```

```{r Analysis 1B: Visualize Dimensional Reduction}
library(patchwork)
library(ggplot2)

# Visualize sample ID
p1 <- dittoDimPlot(spe, var = "sample_id", reduction.use = "UMAP", size = 0.2) +
  scale_colour_manual(values = metadata(spe)$color_vectors$sample_id)
  ggtitle("Sample ID on UMAP")
# p2 <- dittoDimPlot(spe, var = "sample_id", reduction.use = "TSNE", size = 0.2) +
#   scale_colour_manual(values = metadata(spe)$color_vectors$sample_id)
#   ggtitle("Sample ID on UMAP")

# Visualize Indication
p3 <- dittoDimPlot(spe, var = "indication", reduction.use = "UMAP", size = 0.2) +
  scale_colour_manual(values = metadata(spe)$color_vectors$indication)
  ggtitle("Sample ID on UMAP")
# p4 <- dittoDimPlot(spe, var = "indication", reduction.use = "TSNE", size = 0.2) +
#   scale_colour_manual(values = metadata(spe)$color_vectors$indication)
#   ggtitle("Sample ID on UMAP") 
  
# Visualize patient ID
p5 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "UMAP", size = 0.2) +
  scale_colour_manual(values = metadata(spe)$color_vectors$patient_id)
  ggtitle("Sample ID on UMAP")
# p6 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "TSNE", size = 0.2) +
#   scale_colour_manual(values = metadata(spe)$color_vectors$patient_id)
#   ggtitle("Sample ID on UMAP") 

# Plot
#final <- (p1 + p2) / (p3 + p4) / (p5 + p6)
final <- (p1 + p3) / (p5)
ggsave("DimReduction.pdf", final, width = 12, height = 14)

```

```{r Analysis 2A: Batch Correction}
library(batchelor)

# Set seed for reproducibility
set.seed(03061999)

# Calculate fastMNN batch correction
out <- fastMNN(spe, batch = spe$patient_id,
               auto.merge = TRUE,
               subset.row = rowData(spe)$use_channel,
               assay.type = "exprs")

# Check that order of cells is the same
stopifnot(all.equal(colnames(spe), colnames(out)))

# Transfer the correction results to the main spe object
reducedDim(spe, "fastMNN") <- reducedDim(out, "corrected")

# Save RDS object at this point
#saveRDS(spe, "integrated_JaxBatch1.rds")
```

```{r Analysis 2B: Batch Correction QC Assessment}
library(scater)

# Get the batch correction info
merge_info <- metadata(out)$merge.info
merge_info[,c("left", "right", "batch.size")]
merge_info$lost.var

# Set seed for reproducibility
set.seed(03061999)

# Run UMAP on the batch corrected data
spe <- runUMAP(spe, dimred = "fastMNN", name = "UMAP_mnnCorrected")
```

```{r Analysis 2C: Visualizing Batch Corrected UMAP}
library(cowplot)
library(dittoSeq)
library(viridis)

# visualize patient id before and after MNN correction
p1 <- dittoDimPlot(spe, var = "patient_id",
                   reduction.use = "UMAP", size = 0.2) +
  scale_color_manual(values = metadata(spe)$color_vectors$patient_id)+
  ggtitle("Patient ID on UMAP Before Correction")
p2 <- dittoDimPlot(spe, var = "patient_id",
                   reduction.use = "UMAP_mnnCorrected", size = 0.2) +
  scale_color_manual(values = metadata(spe)$color_vectors$patient_id) +
  ggtitle("Patient ID on UMAP After Correction")

# Plot
correctedUMAP <- plot_grid(p1, p2)
ggsave("FastMNN_Corrected_UMAP.pdf", correctedUMAP, width = 14, height = 14)

# visualize sample id after MNN correction
p3 <- dittoDimPlot(spe, var = "sample_id",
                   reduction.use = "UMAP_mnnCorrected", size = 0.2) +
  scale_color_manual(values = metadata(spe)$color_vectors$sample_id) +
  ggtitle("Sample ID on UMAP After Correction")

# visualize indication after MNN correction
p4 <- dittoDimPlot(spe, var = "indication",
                   reduction.use = "UMAP_mnnCorrected", size = 0.2) +
  scale_color_manual(values = metadata(spe)$color_vectors$indication) +
  ggtitle("Indication ID on UMAP After Correction")

# visualize type after MNN correction
p5 <- dittoDimPlot(spe, var = "type",
                   reduction.use = "UMAP_mnnCorrected", size = 0.2) +
  scale_color_manual(values = metadata(spe)$color_vectors$type) +
  ggtitle("Tumor-Type ID on UMAP After Correction")

# Plot as a grid
SampleAndPatient <- plot_grid(p2, p3, p4, p5)
ggsave("Patient_And_Sample_Corrected_UMAP.png", SampleAndPatient, width = 12, height = 8)

```

```{r Analysis 3: Visualizing Each Marker on MNN Corrected UMAP}

# All markers
markers <- rownames(spe)

# Plot each marker on the MNN corrected UMAP
plot_list <- multi_dittoDimPlot(spe, var = markers, reduction.use = "UMAP_mnnCorrected",
                                assay = "exprs", size = 0.2, list.out = TRUE)
plot_list <- lapply(plot_list, function(x) x + scale_color_viridis())
after <- plot_grid(plotlist = plot_list)
ggsave("FastMNN_Corrected_Markers_UMAP.png", after, width = 22, height = 18)

```

```{r Analysis 4A: Kmeans Clustering with 10 centers}
library(bluster)
library(BiocParallel)
library(ggplot2)
library(scran)

# Set seed for reproducibility
set.seed(03061999)

# Cluster the cells, specifying 10 centers
clusters <- clusterCells(spe,
                         use.dimred = "fastMNN",
                         BLUSPARAM = KmeansParam(centers = 10))

# Append clusters to the spe object
spe$k_clusters <- clusters

# Plot the clusters on MNN corrected UMAP
Kmeans_Clusters <- dittoDimPlot(spe, var = "k_clusters",
            reduction.use = "UMAP_mnnCorrected", size = 0.2,
            do.label = TRUE) +
  ggtitle("Kmeans Clusters on UMAP, Integrated Cells")
ggsave("Kmeans_clusters_on_UMAP.pdf", Kmeans_Clusters, width = 10, height = 10)
```

```{r Analysis 4B: Assign colors to KMeans Clusters}

# Assign color vector to KMeans Clusters
KMclusters <- setNames(brewer.pal(length(unique(spe$k_clusters)), name = "PiYG"),
                                  unique(spe$k_clusters))

# Initializ empty list to store cluster color metadata
cluster_colors <- list()

# Append KMclusters color vector to list
cluster_colors$KMclusters <- KMclusters

# Append cluster_colors list to metadata slot of spe
metadata(spe)$cluster_colors <- cluster_colors
```

```{r Analysis 4C: KMeans Clusters Heatmap}

# Sample the cells
cur_cells <- sample(seq_len(ncol(spe)), 2000)

# Plot heatmap of clusters
K_clusterHM <- dittoHeatmap(spe[,cur_cells],
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("k_clusters", "sample_id"),
             annotation_colors = list(KMClusters = metadata(spe)$cluster_colors$KMclusters,
                                      sample_id = metadata(spe)$color_vectors$sample_id))
ggsave("Kmeans_Clusters_Heatmap.pdf", K_clusterHM, width = 8, height = 8)

```

```{r Analysis 5A: Phenotyping}
library(dplyr)

# Assign cluster phenotype labels
phenotypes <- recode(spe$k_clusters,
                           "1" = "Granulocytes",
                           "2" = "Fibroblasts",
                           "3" = "B-cells",
                           "4" = "Monocytes",
                           "5" = "Non-proliferating Epithelia",
                           "6" = "Proilferating Epithelia",
                           "7" = "Mast Cells",
                           "8" = "CAFs",
                           "9" = "T-cells",
                           "10" = "Normal Mesenchymal")
spe$phenotypes <- phenotypes

# Plot MNN corrected UMAP with new cluster labels
phenotypedClusters <- dittoDimPlot(spe, var = "phenotypes",
            reduction.use = "UMAP_mnnCorrected", size = 0.2,
            do.label = TRUE) +
  ggtitle("Phenotyped Clusters on UMAP, Integrated Cells")
ggsave("Phenotyped_mnnCorrected_Clusters.pdf", phenotypedClusters, width = 12, height = 8)

# Plot as heatmap
phenotyped_clusterHM <- dittoHeatmap(spe[,cur_cells],
             genes = rownames(spe),
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("phenotypes",
                          "patient_id",
                          "sample_id",
                          "indication",
                          "type"))
ggsave("Phenotyped_Clusters_Heatmap.pdf", phenotyped_clusterHM, width = 12, height = 8)

```

```{r 5B: Assigning colors to phenotypes clusters}
# Assign colors to each cluster
phenoColors <- setNames(c("#3F1B03", "#F4AD31", "#894F36", "#1C750C", "#EF8ECC", 
                       "#6471E2", "#4DB23B", "grey", "#F4800C", "#BF0A3D"),
                     c("1" = "Granulocytes",
                           "2" = "Fibroblasts",
                           "3" = "B-cells",
                           "4" = "Monocytes",
                           "5" = "Non-proliferating Epithelia",
                           "6" = "Proilferating Epithelia",
                           "7" = "Mast Cells",
                           "8" = "CAFs",
                           "9" = "T-cells",
                           "10" = "Normal Mesenchymal"))
   
# Initialize empty list for cluster colors metadata                  
cluster_colors <- list()

# Append colors to list
cluster_colors$phenotypes <- phenoColors

# Append list to metadata slot
metadata(spe)$cluster_colors <- cluster_colors
```

```{r Analysis 6A: Single Cell Visualization: Heatmaps}
library(scuttle)

# Aggregate acriss cells, calculate the mean expression of each marker
celltype_mean <- aggregateAcrossCells(as(spe, "SingleCellExperiment"),
                                      ids = spe$phenotypes,
                                      statistics = "mean",
                                      use.assay.type = "exprs",
                                      subset.row = rownames(spe))

# Heatmap: no scaling
dittoHeatmap(celltype_mean,
             assay = "exprs",
             cluster_cols = TRUE,
             scale = "none",
             heatmap.colors = viridis(100),
             annot.by = c("phenotypes", "ncells"))

# Max- Scaled
# In max scaling, each feature is scaled to a specific range (min value = 0, max value = 1, and in betwen are scaled linearly)
# Sensitive to outliers because it is based on the minimum and maximum values
# Use when the features have a known, meaningful maximum and minimum.

# Heatmap: min/max scaled
maxScaled <- dittoHeatmap(celltype_mean,
             assay = "exprs",
             cluster_cols = TRUE,
             scaled.to.max = TRUE,
             heatmap.colors.max.scaled = inferno(100),
             annot.by = c("phenotypes", "ncells"),
             annotation_colors = list(ncells = plasma(100)))
ggsave("Max_Scaled_Expression_of_clusters.pdf", maxScaled, width = 10, height = 8)


# Z-score scaled
# Each feature is standardized (has a mean of 0 and a standard deviation of 1. Achieved by subtracting the mean and dividing by the standard deviation)
# Used in situations where the data is assumed to be normal

# Heatmap: Z-scaled
Zscaled <- dittoHeatmap(celltype_mean,
             assay = "exprs",
             cluster_cols = TRUE,
             heatmap.colors.max.scaled = inferno(100),
             annot.by = c("phenotypes", "ncells"),
             annotation_colors = list(ncells = plasma(100)))
ggsave("Z_Scaled_Expression_of_clusters.pdf", Zscaled, width = 10, height = 8)
```


```{r Analysis 6B: Single Cell Visualization: Stacked Barplot}
StackedBar <- dittoBarPlot(spe,
             var = "phenotypes",
             group.by = "sample_id") +
  ggtitle("Cell-type composition of ROIs")

ggsave("StackedBar_Composition_of_ROIs.pdf", StackedBar, width = 10, height = 8)



```

```{r Analysis 7: Project clusters onto the ROIs}
library(gridGraphics)

# Cluster projections
projections <- plotCells(cur_masks,
          object = spe,
          cell_id = "ObjectNumber",
          img_id = "sample_id",
          colour_by = "phenotypes",
          return_plot = TRUE)

# Plot
p1 <- ggdraw(projections$plot, clip = "on")
ggsave("Projected_Clusters.pdf", p1, width =12, height = 8)

```

This loop is too memory intensive...do not run.
```{r , eval = FALSE}
for (img in names(images)) {
  
  norm_img <- cytomapper::normalize(img, inputRange = c(0,0.2))
  norm_mask <- masks[img]

  # Look at the segmentation using a nuclear marker and a few cytoplasmic markers
  plotPixels(image = norm_img,
            mask = norm_mask,
            object = spe,
            cell_id = "ObjectNumber", 
            img_id = "sample_id",
            missing_colour = "white",
            colour_by = c("DNA1","Epcam_ECadherin", "CD3", "CD20"),
            colour = list(DNA1 = c("black", "blue"),
                          Epcam_ECadherin = c("black", "green"),
                          CD3 = c("black", "red"),
                          CD20 = c("black", "orange")),
            image_title = NULL,
            legend = list(colour_by.title.cex = 0.7,
                          colour_by.labels.cex = 0.7))
}
```

```{r Analysis 8: Plot Pixels for Each ROI Manually}
library(gridGraphics)

# Manually go through and make the images for each ROI because your computer cannot handle doing it in a loop.
ROI1 <- images["ROI_008"]
Mask1 <- masks["ROI_008"]
cur_1 <- cytomapper::normalize(ROI1, inputRange = c(0,0.2))

# Look at the segmentation using a nuclear marker and a few cytoplasmic markers
ROIplots <- plotPixels(image = cur_1,
           mask = Mask1,
           object = spe,
           cell_id = "ObjectNumber", 
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA1","Epcam_ECadherin", "CD3", "CD20"),
           colour = list(DNA1 = c("black", "blue"),
                         Epcam_ECadherin = c("black", "green"),
                         CD3 = c("black", "red"),
                         CD20 = c("black", "orange")),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7),
           return_plot = TRUE)

# Plot
ROI <- ggdraw(ROIplots$plot, clip = "on")
ggsave("ROI8.pdf", ROI, width =12, height = 8)
```

From: https://bodenmillergroup.github.io/IMCDataAnalysis/
12.1 Spatial Interaction Graphs
Many spatial analysis approaches either compare the observed versus expected number of cells around a given cell type (point process) or utilize interaction grpahs (spatial object graphs) to estimate clustering or interaction frequencies between cell types.

The `steinbock` framework allows for the construction of these sptial graphs. During image processing, we have constructed a spatial grapsh by expanding the individual cell masks by 4 pixels.

The `imcRtools` package further allows the ad hoc construction of spatial graphs by directly using a `spatialExperiment` or `singleCellExperiment` object while considering the spatial location (centroids) of individual cells. The `buildSpatialGrpah` function allows constructing spatial graphs by detecting the k-nearest neighbors in 2D `(knn)`, by detecting all cells within a given distance to the center cell `(expansion)` and by Delaunay triagnulation `(delaunay)`.

When constructing a knn graph, the number of neighbors `(k)` needs to be set and (optionally) the maximum distance to consider `(max_dist)` can be specified. When constructing a graph via expansion, the distance to expand `(threshold)` needs to be provided. For graphs constructed via Delaunay triangulation, the `max_dist` parameter can be set to avoid unusually large connections at the edge of the image.

```{r Analysis 9A: Building Neighborhood Graphs}
library(imcRtools)

# Build spatial graphs
spe <- buildSpatialGraph(spe, img_id = "sample_id", type = "knn", k = 20)
spe <- buildSpatialGraph(spe, img_id = "sample_id", type = "expansion", threshold = 20)
spe <- buildSpatialGraph(spe, img_id = "sample_id", type = "delaunay", max_dist = 20)
```

```{r Analysis 9B: Plotting Neighborhood Graphs}
library(ggplot2)
library(viridis)
library(gridGraphics)

# Steinbock interaction graph
Spatial <- plotSpatial(spe[,spe$sample_id == "ROI_004"],
            node_color_by = "CD15",
            assay_type = "exprs",
            img_id = "sample_id",
            draw_edges = TRUE,
            colPairName = "delaunay_interaction_graph",
            nodes_first = FALSE,
            node_size_by = "area",
            directed = FALSE,
            edge_color_fix = "grey") +
  scale_size_continuous(range = c(0.1, 2)) +
  ggtitle("ROI4: LOCRC Tumor - CD15 Delaunay Interaction") +
  theme(legend.title = element_text(size = 30),
        legend.text = element_text(size = 20),
        axis.text.x = element_text(size = 30),
        axis.text.y = element_text(size = 30),
        title = element_text(size = 30))
ggsave("CD15_ROI4_Delaunay.pdf", Spatial, width = 14, height = 12)
```

Here's a general guide on interpreting the Delaunay triangulation graph projected onto the ROI:
1.) Spatial organization:
  Triangles: Each triangle in the Delaunay Triangulation graph represents a set of three neighboring cells that are closely related in terms of their protein expression patterns. The edges of these triangles connect cells that are spatially close to each other.

2.) Proximity and Interaction
  Close proximity: Cells connected by edges in the Delaunay triangulation are likely to be in close spatial proximity to each other. This could indicate potential physical interactions or shared micro environments between these cells.
  
3.) Heterogeneity
  Variability in Triangles: The presence of different-sized triangles or irregular patterns may indicate spatial heterogeneity in protein expression within the ROI. This could correspond to distinct cellular neighborhoods or micro environments.
  
4.) Functional Significance
  Protein Co-expression Patterns: Consider the protein expression patterns of cells within each triangle. Are there specific combinations of proteins that tend to co localize spatially? This information can provide insights into potential functional relationships between cells.
  
5.) Biomarker Discovery
  Identification of Key Cells: Cells located at the vertices of triangles or at the center of larger triangles might be of particular interest. These cells may exhibit unique protein expression profiles and could serve as potential biomarkers or indicators of specific cellular states or functions
  





